WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2014-08-12 ms.tex
Reformat for PRX

\documentclass{iopart}
\usepackage{arydshln}
\usepackage{multirow}
\usepackage{amsopn}
\usepackage{iopams}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{bib/aas_macros}
\usepackage[colorlinks]{hyperref}
\usepackage{protosem}
\input{ligo-acronyms/acronyms}

% From http://www.f.kth.se/~ante/latex.php
\setlength{\marginparwidth}{1.2in}
\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%
{\raggedright\footnotesize #1}}

@@

\title[Rapid-response Bayesian sky localization]{WHOOMP! (There It Is) \\
Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients}
\author{Leo Singer and Larry R. Price}
\address{\acs{LIGO} Laboratory, California Institute of Technology, Pasadena, CA 91125, USA}
\ead{\mailto{leo.singer@ligo.org}, \mailto{larryp@caltech.edu}}

@@

@@

In the frequency domain, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}\label{eq:signal-model}
    x_i (\omega) = h_i (\omega; \boldsymbol\theta) + n_i (\omega),
\end{equation}
%
where $h_i (\omega; \boldsymbol\theta)$ is the \ac{GW} signal given a parameter vector $\boldsymbol\theta$ that describes the \ac{GW} source, and $n_i (\omega)$ is that detector's \ac{WSS} Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[\left|n_i(\omega)\right|^2\right] + E\left[\left|n_i(-\omega)\right|^2\right] = 2 E\left[\left|n_i(\omega)\right|^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf x (\omega) \equiv \{x_i (\omega)\}_i$.

The likelihood, or the probability of obtaining the observation $\mathbf x$ conditioned on the value of $\boldsymbol\theta$, is a product of Gaussian distributions:
%
\begin{equation}\label{eq:gaussian-likelihood}
    \mathcal{L}(\mathbf x; \boldsymbol\theta) = \prod_i p(x_i | \boldsymbol\theta)
        \propto \exp \left[
        - \frac{1}{2} \sum_i \int_0^\infty \frac{\left|x_i (\omega)
            - h_i(\omega; \boldsymbol\theta) \right|^2}{S_i(\omega)} \, d\omega
    \right].
\end{equation}

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \boldsymbol\theta = \begin{array}{l@{}l}
            \left[
            \begin{array}{c}
                \alpha \\
                \delta \\
                r \\
                t_\oplus \\
                \iota \\
                \psi \\
                \phi_c \\
                \hdashline[1pt/1pt]
                m_1 \\
                m_2 \\
                \mathbf S_1 \\
                \mathbf S_2
            \end{array}
            \right] &
            \begin{array}{l}
                \left.
                \begin{array}{p{4cm}}
                    right ascension \\
                    declination \\
                    distance \\
                    arrival time at geocenter \\
                    inclination angle \\
                    polarization angle \\
                    coalescence phase \\
                    \hdashline[1pt/1pt]
                \end{array}
                \quad \right\} \textnormal{extrinsic parameters}, \boldsymbol\theta_\mathrm{ex} \\
                \left.
                \begin{array}{p{4cm}}
                    first component's mass \\
                    second component's mass \\
                    first component's spin \\
                    second component's spin
                \end{array}
                \quad \right\} \textnormal{intrinsic parameters}, \boldsymbol\theta_\mathrm{in}
            \end{array}
        \end{array}
\end{equation}
%
This list of parameters involves some simplifying assumptions. Eccentricity is omitted: although it does play a major role in the evolution and waveforms of \ac{NSBH} and \ac{BBH} sources formed by dynamical capture~\cite{PhysRevD.87.043004}, \ac{BNS} systems formed by binary stellar evolution should circularize due to tidal interaction~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$.

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $h_0$ and $h_{\pi/2}$. For nonprecessing systems, $h_0$ and $h_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. If $h_0$ and $h_{\pi/2}$ are Fourier transforms of real functions, then $h_0(\omega) = h_0^*(-\omega)$ and $h_{\pi/2}(\omega) = h_{\pi/2}^*(-\omega)$, and we can write (assuming an arbitrary phase convention)
%
\begin{equation}
    h_{\pi/2}(\omega) = h_0(\omega) \cdot
    \cases{
        -i & if $\omega \geq 0$ \\
        i & if $\omega < 0$
    }.
\end{equation}
%
Then, we can write the signal model in a way that isolates all dependence on the extrinsic parameters $\boldsymbol\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\boldsymbol\theta_\mathrm{in}$ into the basis waveform, by taking the Fourier transform of Equation~(2.8) of \cite{PhysRevD.83.084002}:
%
\begin{equation}\label{eq:full-signal-model}
    h_i(\omega; \boldsymbol\theta) = \e^{-i \omega t_\oplus}
    \frac{r_{1,i}}{r}
    \e^{2 i \phi_c}
    \left[
    \frac{1}{2} \left(1 + \cos^2 \iota\right) \Re \left\{z\right\} - i
    \left(\cos\iota\right) \Im z
    \right]
    h(\omega; \boldsymbol\theta_\mathrm{in})
\end{equation}
%
for $\omega \geq 0$, where
%
\begin{equation}
    z = e^{-2 i \psi} \left(
    F_{+,i}(\alpha, \delta, t_\oplus) +
    i F_{\times,i}(\alpha, \delta, t_\oplus)
    \right).
\end{equation}
%
The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, defined such that $0 \leq {F_{+,i}}^2 + {F_{\times,i}} \leq 1$. They depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event, and are derived in~\cite{PhysRevD.63.042003}. The quantity $r_{1,i}$ is a fiducial distance at which detector $i$ would register SNR=1 for an optimally oriented binary (face\nobreakdashes-on, and in a direction perpendicular to the interferometer's arms),
%
\marginpar{I seem to have dropped a factor of 2 here...}%
%
\begin{equation}\label{eq:horizon}
r_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|h(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,d\omega.
\end{equation}

More succinctly, we can write the signal received by detector $i$ in terms of observable extrinsic parameters $\boldsymbol\theta_i = (\rho_i, \gamma_i, \tau_i)$, the amplitude $\rho_i$, phase $\gamma_i$, and time delay $\tau_i$ on arrival at detector $i$,
%
\begin{equation}\label{eq:signal-model}
    h_i (\omega; \boldsymbol\theta_i, \boldsymbol\theta_\mathrm{in}) = h_i (\omega; \rho_i, \gamma_i, \tau_i, \boldsymbol\theta_\mathrm{in}) = \frac{\rho_i}{\sigma_i} e^{i (\gamma_i - \omega \tau_i)} h(\omega; \boldsymbol\theta_\mathrm{in}).
\end{equation}

The prevailing technique for detection of \acp{GW} from \acp{CBC} is to realize a \ac{MLE} from the likelihood in \Eref{eq:gaussian-likelihood} and the signal model in \Eref{eq:signal-model}. Concretely, this results in a bank of matched filters, or the cross-correlation between a template waveform and the incoming data stream,
%
\begin{equation}
y_i(\tau_i;\boldsymbol\theta_\mathrm{in}) = \frac{1}{\sigma_i (\boldsymbol\theta_\mathrm{in})} \int_0^\infty \frac{h^*(\omega; \boldsymbol\theta_\mathrm{in}) x_i(\omega) e^{i \omega \tau_i}}{S_i(\omega)} \,d\omega.
\end{equation}
%
The \ac{ML} point estimates of the signal parameters, $\mathrm{MLE}(\mathbf{x}) = \{\{ \hat{\boldsymbol\theta}_i \}_i, \hat{\boldsymbol\theta}_\mathrm{in}\} = \{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\boldsymbol\theta}_\mathrm{in}\}$, are given by
%
\begin{eqnarray}
    \hat{\boldsymbol\theta}_\mathrm{in}, \{\hat\tau_i\}_i
        &=& \argmax_{\boldsymbol\theta_\mathrm{in}, \{\hat\tau_i\}_i}
        \sum_i \left| y_i\left(\tau_i;
        \boldsymbol\theta_\mathrm{in}\right) \right|^2, \\
    \hat\rho_i &=& \left| y_i\left(\hat\tau_i;
        \hat{\boldsymbol\theta}_\mathrm{in}\right) \right|, \\
    \hat\gamma_i &=& \arg y_i\left(\hat\tau_i;
        \hat{\boldsymbol\theta}_\mathrm{in}\right).
\end{eqnarray}
%
A detection candidate consists of $\{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\boldsymbol\theta}_\mathrm{in}\}$. There are various ways to characterize the significance of a detection candidate, but in Gaussian noise the combined SNR $\hat\rho_\mathrm{c}$ suffices,
%
\marginpar{What is the least that I can get away with saying about false alarm rate estimation and likelihood ratios?}%
%
\begin{equation}
    \hat\rho_\mathrm{c} = \sqrt{\sum_i {\hat\rho_i}^2}.
\end{equation}

@@

In the Bayesian framework, the parameters are inferred from the data by forming the posterior distribution, $p(\boldsymbol\theta|\mathbf x)$, which describes the probability of the parameters given the observations. Bayes' rule relates the likelihood $p(\mathbf x|\boldsymbol\theta)$ to the posterior $p(\boldsymbol\theta|\mathbf x)$,
%
\begin{equation}\label{bayes}
p(\boldsymbol\theta|\mathbf x) = \frac{p(\mathbf x|\boldsymbol\theta) p(\boldsymbol\theta)}{p(\mathbf x)},
\end{equation}
%
introducing the prior distribution $p(\boldsymbol\theta)$ which encapsulates previous information about the paramters (for example, arising from earlier observations or from known physical bounds on the parameters), and the evidence $p(\mathbf x)$ which can be thought of as a normalization factor or as describing the parsimoniousness of the model.

@@

In Bayesian inference, although it is often easy to write down the likelihood or even the full posterior in closed form, usually one is interested in only a subset $\boldsymbol\beta$ of all of the model's parameters, the others $\boldsymbol\lambda$ being nuisance parameters. In this case, we integrate away the nuisance parameters, forming the marginal posterior
%
\begin{equation}\label{eq:marginal-posterior}
    p(\boldsymbol\beta|\mathbf x) = \int \frac{p(\mathbf x|\boldsymbol\beta,\boldsymbol\lambda) p(\boldsymbol\beta,\boldsymbol\lambda)}{p(\mathbf x)} \,d\boldsymbol\lambda
\end{equation}
%
with $\boldsymbol\theta = (\boldsymbol\beta, \boldsymbol\lambda)$. For instance, for the purpose of locating a \ac{GW} source on the sky, all parameters but $(\alpha, \delta)$ are nuisance parameters.

@@

For the purpose of rapid sky localization, we assume that we do not have access to the \ac{GW} data $\mathbf{x}$ itself, and that our only contact with it is through the ML parameter estimates $\{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\boldsymbol\theta}_\mathrm{in}\}$. Although this is a significant departure from conventional GW parameter estimation techniques, we can still apply the full Bayesian machinery of \Eref{eq:marginal-posterior} to compute a posterior distribution for the sky location.

@@

To treat the ML estimates as measurements in and of themselves, we need to work out their likelihood, marginalized over the GW signal $\mathbf{x}$,
%
\begin{equation}\label{eq:detection-candidate-likelihood}
    p\left(\{\hat{\boldsymbol\theta}_i\}_i,
        \hat{\boldsymbol\theta}_\mathrm{in}
    \middle| \boldsymbol\theta\right)
    \propto \int\limits_{\mathclap{\mathbf{x} | \mathrm{MLE}(\mathbf{x}) =
        \{\{\hat{\boldsymbol\theta}_i\}_i,
        \hat{\boldsymbol\theta}_\mathrm{in}\}}}
    p(\mathbf{x} | \boldsymbol\theta) \, p(\boldsymbol\theta)
    \, d\mathbf{x}
\end{equation}
%
The integral above could be written in terms of the matched filter signals $\left\{y_i\right\}_i$, such that $y_i\left(\hat\tau_i, \hat{\boldsymbol\theta}_\mathrm{in}\right) = \hat\rho_i e^{i \hat\gamma_i}$ and $\forall t$, $\left|y_i(t, \boldsymbol\theta_\mathrm{in})\right| \leq \hat\rho_i$. The combination of the nontrivial integration limits and the Fourier transform in the matched filter seem like a problem for direct evaluation.

However, even without directly evaluating the distribution of the \ac{ML} estimator, we know many of its properties. The \ac{CRLB} gives its covariance in the asymptotic limit of high \ac{SNR}. The \ac{CRLB} has been widely applied in \ac{GW} data analysis \cite{fairhurst:2009}. We will momentarily consider the likelihood for a single detector:
%
\begin{equation}\label{eq:gaussian-likelihood-spa}
    \mathcal{L}\left(x_i; \rho_i, \gamma_i, \tau_i,
        \boldsymbol\theta_\mathrm{in}\right)
    \propto \exp \left[
        - \frac{1}{2} \int_0^\infty \frac{\left|x_i (\omega)
            - h_i\left(x_i; \rho_i, \gamma_i, \tau_i,
                \boldsymbol\theta_\mathrm{in}\right)
        \right|^2}{S_i(\omega)} \, d\omega
    \right],
\end{equation}
%
with $h_i(\omega; \rho_i, \gamma_i, \tau_i, \boldsymbol\theta_\mathrm{in})$ given by \Eref{eq:signal-model}.

>>>>

\documentclass[amsmath,amssymb,aps,prx,reprint,nopreprintnumbers,nofootinbib]{revtex4-1}
\usepackage{multirow}
\usepackage{bigdelim}
\usepackage{multirow}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{bib/aas_macros}
\usepackage[colorlinks]{hyperref}
\usepackage{protosem}
\input{ligo-acronyms/acronyms}

% From http://www.f.kth.se/~ante/latex.php
\setlength{\marginparwidth}{0.5in}
\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\scriptsize #1]%
{\raggedright\scriptsize #1}}

@@

\title[Rapid-response Bayesian sky localization]{WHOOMP! (There It Is) \\
Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients}
\author{Leo P. Singer}
\email{leo.singer@ligo.org}
\author{Larry R. Price}
\email{larryp@caltech.edu}
\affiliation{\acs{LIGO} Laboratory, California Institute of Technology, Pasadena, CA 91125, USA}

@@

\maketitle

@@

In the frequency domain, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}\label{eq:signal-model}
    x_i (\omega) = h_i (\omega; \bm\theta) + n_i (\omega),
\end{equation}
%
where $h_i (\omega; \bm\theta)$ is the \ac{GW} signal given a parameter vector $\bm\theta$ that describes the \ac{GW} source, and $n_i (\omega)$ is that detector's \ac{WSS} Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[\left|n_i(\omega)\right|^2\right] + E\left[\left|n_i(-\omega)\right|^2\right] = 2 E\left[\left|n_i(\omega)\right|^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf x (\omega) \equiv \{x_i (\omega)\}_i$.

The likelihood, or the probability of obtaining the observation $\mathbf x$ conditioned on the value of $\bm\theta$, is a product of Gaussian distributions:
%
\begin{multline}\label{eq:gaussian-likelihood}
    \mathcal{L}(\mathbf x; \bm\theta) = \prod_i p(x_i | \bm\theta)
        \\ \propto \exp \left[
        - \frac{1}{2} \sum_i \int_0^\infty \frac{\left|x_i (\omega)
            - h_i(\omega; \bm\theta) \right|^2}{S_i(\omega)} \, d\omega
    \right].
\end{multline}

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \bm\theta = \left[
    \begin{array}{cl@{\quad}p{4cm}lp{3.75cm}}
        \alpha & \rdelim]{11}{0mm} & right ascension & \rdelim\}{7}{1mm} & \multirow{7}{2cm}{extrinsic parameters, $\bm\theta_\mathrm{ex}$} \\
        \delta && declination & \\
        r && distance & \\
        t_\oplus && arrival time at geocenter & \\
        \iota && inclination angle & \\
        \psi && polarization angle & \\
        \phi_c && coalescence phase & \\
        \cline{1-1}\cline{3-3}
        m_1 && first component's mass & \rdelim\}{4}{1mm} & \multirow{4}{2cm}{intrinsic parameters, $\bm\theta_\mathrm{in}$.}\\
        m_2 && second component's mass & \\
        \mathbf S_1 && first component's spin & \\
        \mathbf S_2 && second component's spin & \\
    \end{array}\right.
\end{equation}
%
This list of parameters involves some simplifying assumptions. Eccentricity is omitted: although it does play a major role in the evolution and waveforms of \ac{NSBH} and \ac{BBH} sources formed by dynamical capture~\cite{PhysRevD.87.043004}, \ac{BNS} systems formed by binary stellar evolution should circularize due to tidal interaction~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$.

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $h_0$ and $h_{\pi/2}$. For nonprecessing systems, $h_0$ and $h_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. If $h_0$ and $h_{\pi/2}$ are Fourier transforms of real functions, then $h_0(\omega) = h_0^*(-\omega)$ and $h_{\pi/2}(\omega) = h_{\pi/2}^*(-\omega)$, and we can write (assuming an arbitrary phase convention)
%
\begin{equation}
    h_{\pi/2}(\omega) = h_0(\omega) \cdot
    \begin{cases}
        -i & \text{if } \omega \geq 0 \\
        i & \text{if } \omega < 0
    \end{cases}.
\end{equation}
%
Then, we can write the signal model in a way that isolates all dependence on the extrinsic parameters $\bm\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\bm\theta_\mathrm{in}$ into the basis waveform, by taking the Fourier transform of Equation~(2.8) of \cite{PhysRevD.83.084002}:
%
\begin{multline}\label{eq:full-signal-model}
    h_i(\omega; \bm\theta) = e^{-i \omega t_\oplus}
    \frac{r_{1,i}}{r}
    e^{2 i \phi_c} \\
    \left[
    \frac{1}{2} \left(1 + \cos^2 \iota\right) \Re \left\{z\right\} - i
    \left(\cos\iota\right) \Im z
    \right]
    h(\omega; \bm\theta_\mathrm{in})
\end{multline}
%
for $\omega \geq 0$, where
%
\begin{equation}
    z = e^{-2 i \psi} \left(
    F_{+,i}(\alpha, \delta, t_\oplus) +
    i F_{\times,i}(\alpha, \delta, t_\oplus)
    \right).
\end{equation}
%
The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, defined such that $0 \leq {F_{+,i}}^2 + {F_{\times,i}} \leq 1$. They depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event, and are derived in~\cite{PhysRevD.63.042003}. The quantity $r_{1,i}$ is a fiducial distance at which detector $i$ would register SNR=1 for an optimally oriented binary (face\nobreakdashes-on, and in a direction perpendicular to the interferometer's arms),
%
\marginpar{I seem to have dropped a factor of 2 here...}%
%
\begin{equation}\label{eq:horizon}
r_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|h(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,d\omega.
\end{equation}

More succinctly, we can write the signal received by detector $i$ in terms of observable extrinsic parameters $\bm\theta_i = (\rho_i, \gamma_i, \tau_i)$, the amplitude $\rho_i$, phase $\gamma_i$, and time delay $\tau_i$ on arrival at detector $i$,
%
\begin{equation}\label{eq:signal-model}
    h_i (\omega; \bm\theta_i, \bm\theta_\mathrm{in}) = h_i (\omega; \rho_i, \gamma_i, \tau_i, \bm\theta_\mathrm{in}) = \frac{\rho_i}{\sigma_i} e^{i (\gamma_i - \omega \tau_i)} h(\omega; \bm\theta_\mathrm{in}).
\end{equation}

The prevailing technique for detection of \acp{GW} from \acp{CBC} is to realize a \ac{MLE} from the likelihood in Equation~(\ref{eq:gaussian-likelihood}) and the signal model in Equation~(\ref{eq:signal-model}). Concretely, this results in a bank of matched filters, or the cross-correlation between a template waveform and the incoming data stream,
%
\begin{equation}
y_i(\tau_i;\bm\theta_\mathrm{in}) = \frac{1}{\sigma_i (\bm\theta_\mathrm{in})} \int_0^\infty \frac{h^*(\omega; \bm\theta_\mathrm{in}) x_i(\omega) e^{i \omega \tau_i}}{S_i(\omega)} \,d\omega.
\end{equation}
%
The \ac{ML} point estimates of the signal parameters, $\mathrm{MLE}(\mathbf{x}) = \{\{ \hat{\bm\theta}_i \}_i, \hat{\bm\theta}_\mathrm{in}\} = \{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\bm\theta}_\mathrm{in}\}$, are given by
%
\begin{eqnarray}
    \hat{\bm\theta}_\mathrm{in}, \{\hat\tau_i\}_i
        &=& \argmax_{\bm\theta_\mathrm{in}, \{\hat\tau_i\}_i}
        \sum_i \left| y_i\left(\tau_i;
        \bm\theta_\mathrm{in}\right) \right|^2, \\
    \hat\rho_i &=& \left| y_i\left(\hat\tau_i;
        \hat{\bm\theta}_\mathrm{in}\right) \right|, \\
    \hat\gamma_i &=& \arg y_i\left(\hat\tau_i;
        \hat{\bm\theta}_\mathrm{in}\right).
\end{eqnarray}
%
A detection candidate consists of $\{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\bm\theta}_\mathrm{in}\}$. There are various ways to characterize the significance of a detection candidate, but in Gaussian noise the combined SNR $\hat\rho_\mathrm{c}$ suffices,
%
\marginpar{What is the least that I can get away with saying about false alarm rate estimation and likelihood ratios?}%
%
\begin{equation}
    \hat\rho_\mathrm{c} = \sqrt{\sum_i {\hat\rho_i}^2}.
\end{equation}

@@

In the Bayesian framework, the parameters are inferred from the data by forming the posterior distribution, $p(\bm\theta|\mathbf x)$, which describes the probability of the parameters given the observations. Bayes' rule relates the likelihood $p(\mathbf x|\bm\theta)$ to the posterior $p(\bm\theta|\mathbf x)$,
%
\begin{equation}\label{bayes}
p(\bm\theta|\mathbf x) = \frac{p(\mathbf x|\bm\theta) p(\bm\theta)}{p(\mathbf x)},
\end{equation}
%
introducing the prior distribution $p(\bm\theta)$ which encapsulates previous information about the paramters (for example, arising from earlier observations or from known physical bounds on the parameters), and the evidence $p(\mathbf x)$ which can be thought of as a normalization factor or as describing the parsimoniousness of the model.

@@

In Bayesian inference, although it is often easy to write down the likelihood or even the full posterior in closed form, usually one is interested in only a subset $\bm\beta$ of all of the model's parameters, the others $\bm\lambda$ being nuisance parameters. In this case, we integrate away the nuisance parameters, forming the marginal posterior
%
\begin{equation}\label{eq:marginal-posterior}
    p(\bm\beta|\mathbf x) = \int \frac{p(\mathbf x|\bm\beta,\bm\lambda) p(\bm\beta,\bm\lambda)}{p(\mathbf x)} \,d\bm\lambda
\end{equation}
%
with $\bm\theta = (\bm\beta, \bm\lambda)$. For instance, for the purpose of locating a \ac{GW} source on the sky, all parameters but $(\alpha, \delta)$ are nuisance parameters.

@@

For the purpose of rapid sky localization, we assume that we do not have access to the \ac{GW} data $\mathbf{x}$ itself, and that our only contact with it is through the ML parameter estimates $\{\left\{ \hat\rho_i, \hat\gamma_i, \hat\tau_i \right\}_i, \hat{\bm\theta}_\mathrm{in}\}$. Although this is a significant departure from conventional GW parameter estimation techniques, we can still apply the full Bayesian machinery of Equation~(\ref{eq:marginal-posterior}) to compute a posterior distribution for the sky location.

@@

To treat the ML estimates as measurements in and of themselves, we need to work out their likelihood, marginalized over the GW signal $\mathbf{x}$,
%
\begin{equation}\label{eq:detection-candidate-likelihood}
    p\left(\{\hat{\bm\theta}_i\}_i,
        \hat{\bm\theta}_\mathrm{in}
    \middle| \bm\theta\right)
    \propto \int\limits_{\mathclap{\mathbf{x} | \mathrm{MLE}(\mathbf{x}) =
        \{\{\hat{\bm\theta}_i\}_i,
        \hat{\bm\theta}_\mathrm{in}\}}}
    p(\mathbf{x} | \bm\theta) \, p(\bm\theta)
    \, d\mathbf{x}
\end{equation}
%
The integral above could be written in terms of the matched filter signals $\left\{y_i\right\}_i$, such that $y_i\left(\hat\tau_i, \hat{\bm\theta}_\mathrm{in}\right) = \hat\rho_i e^{i \hat\gamma_i}$ and $\forall t$, $\left|y_i(t, \bm\theta_\mathrm{in})\right| \leq \hat\rho_i$. The combination of the nontrivial integration limits and the Fourier transform in the matched filter seem like a problem for direct evaluation.

However, even without directly evaluating the distribution of the \ac{ML} estimator, we know many of its properties. The \ac{CRLB} gives its covariance in the asymptotic limit of high \ac{SNR}. The \ac{CRLB} has been widely applied in \ac{GW} data analysis \cite{fairhurst:2009}. We will momentarily consider the likelihood for a single detector:
%
\begin{multline}\label{eq:gaussian-likelihood-spa}
    \mathcal{L}\left(x_i; \rho_i, \gamma_i, \tau_i,
        \bm\theta_\mathrm{in}\right) \\
    \propto \exp \left[
        - \frac{1}{2} \int_0^\infty \frac{\left|x_i (\omega)
            - h_i\left(x_i; \rho_i, \gamma_i, \tau_i,
                \bm\theta_\mathrm{in}\right)
        \right|^2}{S_i(\omega)} \, d\omega
    \right],
\end{multline}
%
with $h_i(\omega; \rho_i, \gamma_i, \tau_i, \bm\theta_\mathrm{in})$ given by Equation~(\ref{eq:signal-model}).

The Fisher information matrix for a measurement $x$ described by the unknown parameter vector $\bm{\theta}$ is
%
\begin{equation}\label{eq:general-fisher-matrix}
    \mathcal{I}_{jk} = \mathrm{E} \, \left[
        -\frac{\partial^2 \log
            \mathcal{L}(x_i ; \bm\theta)}
            {\partial \theta_j \theta_k}
    \middle| \bm\theta
    \right].
\end{equation}
%
In other words, it is the expectation value, condition on the true parameter values, of the Jacobian of the log likelihood. This equation involves double derivatives, but substantial simplification is possible when---as in this case---the likelihood is Gaussian:
%
\begin{equation}\label{eq:gaussian-fisher-matrix}
    \mathcal{I}_{jk} = \int_0^\infty \Re \left[
        \left(\frac{\partial h_i}{\partial \theta_j}\right)^*
        \left(\frac{\partial h_i}{\partial \theta_k}\right)
    \right] \frac{1}{S_i(\omega)} \, d\omega.
\end{equation}
%
This form is more useful because it involves only first derivatives, and of the signal $h (\omega)$ rather than the entire observation $x (\omega)$. In terms of the $k$th \ac{SNR}-weighted moment of angular frequency,
%
\begin{equation}\label{eq:angular-frequency-moments}
    {\overline{\omega^k}}_i =
        \left[ \int_0^\infty \frac{|h (\omega)|^2}{S_i(\omega)} \omega^k \, d\omega \right]
        \left[ \int_0^\infty \frac{|h (\omega)|^2}{S_i(\omega)} \, d\omega \right]^{-1},
\end{equation}
%
the Fisher matrix for the signal in the $i$th detector is
%
\begin{equation}
    \mathcal{I}_i = \left(
        \begin{array}{cc}
            \mathcal{I}_{\bm\theta_i,\bm\theta_i} &
            \mathcal{I}_{\bm\theta_i,\bm\theta_\mathrm{in}} \\
            {\mathcal{I}_{\bm\theta_i,\bm\theta_\mathrm{in}}}^\intercal &
            {\rho_i}^2 \mathcal{I}_{\bm\theta_\mathrm{in},\bm\theta_\mathrm{in}}
        \end{array}
    \right)
\end{equation}
%
where
%
\begin{equation}\label{eq:fisher-matrix}
    \mathcal{I}_{\bm\theta_i,\bm\theta_i} = \bordermatrix{
        ~ & \rho_i & \gamma_i & \tau_i \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2 {\overline{\omega}}_i \cr
        \tau_i & 0 & -{\rho_i}^2 {\overline{\omega}}_i & {\rho_i}^2 {\overline{\omega^2}}_i
    }.
\end{equation}
%
(See also \cite{Grover:2013}.) To lowest post\nobreakdashes-Newtonian order, the intrinsic parameters affect the phase evolution but not the amplitude evolution, so for BNS systems we can assume $\mathcal{I}_{\rho_i,\bm\theta_\mathrm{in}} \approx 0$.
%
\marginpar{Show this?}%
%
When we form the Fisher matrix for the whole network, if the detectors all have the same noise \acp{PSD}, $S_1(\omega) = S_2(\omega) = \cdots = S_n(\omega) \equiv S(\omega)$, then we can change variables from phases and times on arrival $\left\{\gamma_i, \tau_i\right\}_i$ to appropriate differences between them, in order to cancel out all off-diagonal terms between the intrinsic and extrinsic parameters. Because all existing ground-based GW detectors have similar bandpasses, we restrict all of our further attention to just the extrinsic part of the Fisher matrix, $\mathcal{I}_{\bm\theta_i,\bm\theta_i}$.
